Data Migration under a Changing Schema in Ampersand

### 2023-10-02 2022Migration/articleMigrationFACS.tex
refine some formulations (minor)

\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be upgraded at an ever-increasing pace to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly upgrades of existing systems.
   The need to do a data migration arises only if the schema of a new system differs from the existing one.
   Let us call that a "schema changing data migration" (SCDM).
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   We can expect more SCDMs as the deployment frequency increases.

   For good reasons, most development teams will try to avoid schema changes at all costs, fearing the risk and effort of the data migration.
   During the lifetime of a system, SCDMs cannot always be avoided, however.
   It is this problem that we address in this paper.
   Data migration for other purposes has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
   If the schema is not available at compile time, the work will have to be done at runtime.
   This leads to versioned storage and an overhead in performance.
   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
   The prototypical use case for that is to frequently release upgrades of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

   To analyze data migrations with changing schemas, we need to define the notion of schema and how it constrains the data.
   To facilitate a formal analysis of the situation,
   we define an information system (section~\ref{sct:Information Systems}) as a pair of a schema and a dataset,
   so that we have the schema explicitly at our disposal.
   The dataset is separate, so we may consider the analysis of just a schema as a compile-time activity.
   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and rules.
   Relations are carriers of the data and concepts are needed to define them.
   Rules constrain the data to impose semantics on the data.
   These constraints must be satisfied all the time, so we call them {\em invariant}.
   For experimental validation we have used the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
   because its syntax and semantics correspond directly to these definitions.
   Ampersand has a notion of rule, which implements the invariants on the dataset.
   In this way, prototypes built in Ampersand have helped to refine the theory.

   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, invariants that change yield extra work for software engineers to write the transform code,
   for which ETL tools typically provide little support.
   If automation can improve that, the frequency of SCDMs might well increase.
   As a consequence, users will accept downtime for the sake of migration much less.
   So, our research aims for zero-downtime as perceived by end-users.
   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean up data.
   Some of that must be done before the migration; some can wait till after the migration.
   We can define part of the data quality as satisfying semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   We can actually use invariants to help find and correct data pollution.
   We cannot use always use invariants, however.
   An example is when a person has deliberately specified a false name, without violating any of the existing invariants.

   To make the case for zero-downtime, we must distinguish between two types of constraint:
\begin{enumerate}
\item Invariant\\
   An \define{invariant} is a constraint that is always true in a system.
   This is the classical database transaction.
   Violation of an invariant means it has to be restored without the outside world noticing.
   This requires either human intervention or an automated procedure.
   If it is restored, the system commits to the new state, which satisfies the constraint.
   If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
   In both cases, the invariant remains true.
\item Business constraint\\
   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   We consider business constraints not to be invariants because users can violate them.
\end{enumerate}
   Suppose some invariant of the desired system does not apply to the existing system,
   and suppose it would be violated if applied to the data of the existing system.
   However, spinning up the desired system on the data set of the existing system requires
   that every constraint of the desired system is satisfied.
   That is a problem.
   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the process.
   So, keeping the existing system until the invariants of the desired system are satisfied is not an option.
   To solve this problem, we implement every invariant in the migration system initially as a business constraint,
   letting users restore invariance while the new system is running.
   For this purpose, we must define an intermediate system, the \define{migration system},
   which implements the invariants of the desired system as business constraints.
   This migration system ensures that every violation that is fixed will not return as a violation.
   In this way, the business constraint gradually turns into an invariant.
   Since the number of violations is finite, it is up to the business to resolve these violations.
   New violations that occur are under control of the desired system,
   so the business experiences the migration system as the desired system.
   In this way, our research realizes zero-downtime.

   Summarizing, the following requirements apply to SCDMs
\begin{enumerate}
\item users must experience zero-downtime, to enable more frequent SCDMs.
\item users must be able to fix business constraints, so they can be used to spin up a migration system and deliver zero-downtime.
\item the number of violations that users must fix is finite and decreases (monotonically) over time.
\end{enumerate}
   An information system that is defined by constraints on a data set
   can signal certain forms of data pollution by adding the proper constraints.
   The mechanism of business constraints can help to clean that pollution.

>>>>

\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be upgraded at an ever-increasing pace to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly upgrades of existing systems.
   If an upgrade does not change the schema, we don't need a data migration.
   Yet, schema changes cannot always be avoided, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   So, development teams will try to avoid schema changes at all costs, fearing the risk and effort of the data migration,
   and for a good reason.

   Data migration for other purposes than schema change has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
   If the schema is not available at compile time, the work will have to be done at runtime.
   This leads to versioned storage and an overhead in performance.
   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
   The prototypical use case for that is to frequently release upgrades of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

   To analyze data migrations with changing schemas, we need to define the notion of schema and how it constrains the data.
   To facilitate a formal analysis of the situation,
   we define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a dataset,
   so that we have the schema explicitly at our disposal.
   The dataset is separate, so we may consider the analysis of just a schema as a compile-time activity.
   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and rules.
   Relations are carriers of the data and concepts are needed to define them.
   Rules constrain the data to impose semantics on the data.
   These constraints must be satisfied all the time, so we call them {\em invariant}.
   For experimental validation we have used the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
   because its syntax and semantics correspond directly to these definitions.
   Ampersand has a notion of rule, which implements the invariants on the dataset.
   In this way, prototypes built in Ampersand have helped to refine the theory.

   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, invariants that change yield extra work for software engineers to write the transform code,
   for which ETL tools typically provide little support.
   The pressure on increasing deploy times calls for further automation of SCDMs.
   Hence, we can expect more frequent SCDMs, and downtime for the sake of migration becomes less acceptable.
   So, our research aims for zero-downtime.
   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean up data.
   Some of that must be done before the migration; some can wait till after the migration.
   We can define part of the data quality as satisfying semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   An information system that is defined by constraints on a data set
   can signal this type of data pollution by adding such constraints.
   Some forms of data pollution cannot be detected in this way, however.
   An example is when a person has deliberately specified a false name without violating any conceivable constraint.

   To make the case for zero-downtime, we must distinguish between two types of constraint:
\begin{enumerate}
\item Invariant\\
   An \define{invariant} is a constraint that is always true in a system.
   This corresponds to the classical database transaction.
   Violation of an invariant means it has to be restored without the outside world noticing.
   This requires either human intervention or an automated procedure.
   If it is restored, the system commits to the new state, which satisfies the constraint.
   If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
   In both cases, the invariant remains true.
\item Business constraint\\
   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   We consider business constraints not to be invariants because users can violate them.
   They are not true all the time.
\end{enumerate}
   Suppose some invariant, $u$, of the {\em desired system} does not apply to the {\em existing system},
   and suppose it would be violated if applied to the data of the existing system.
   Now, when we spin up the desired system on the data set of the existing system, we have a problem.
   Every invariant of the desired system must be satisfied but $u$ isn't.
   So we cannot spin it up.
   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the process.
   To solve this problem, we implement every invariant in the migration system initially as a business constraint,
   letting users restore invariance while the new system is running.
   For this purpose, we must define an intermediate system, the \define{migration system},
   which implements the invariants of the desired system as business constraints.
   This migration system ensures that every violation that is fixed will not return as a violation.
   In this way, the business constraint gradually turns into an invariant.
   Since the number of violations is finite, it is up to the business to resolve these violations.
   New violations that occur are under control of the desired system,
   so the business experiences the migration system as the desired system.
   In this way, the migration system bridges the gap and users get a zero-downtime SCDM.

   Summarizing, the following requirements apply to SCDMs
\begin{enumerate}
\item users must experience zero-downtime, to enable more frequent SCDMs.
\item users must be able to fix business constraints, so we need an intermediate ``migration system'' to deliver zero-downtime.
\item the number of violations that users must fix is finite and decreases (monotonically) over time,
   to ensure that the migration system can be removed after the migration is done.
\end{enumerate}
